---
title: "Simulating Data for Mars <br> `r emo::ji('rocket')`"
author: "S. Mason Garrison"
output:
  xaringan::moon_reader:
    css: "../slides.css"
    lib_dir: libs
    nature:
      ratio: "16:9"
      highlightLines: true
      highlightStyle: solarized-light
      countIncrementalSlides: false
      slideNumberFormat: ""
---



```{r child = "../setup.Rmd"}
```

```{r packages, echo = FALSE, message=FALSE, warning=FALSE}
library(tidyverse)
library(MASS)
library(emo)
library(viridis)
library(knitr)
# Remember to compile
# xaringan::inf_mr(cast_from = "..")
```

class: middle

# Learning Goals

---

## Learning Goals

By the end of this session, you will be able to...

- Construct simulated datasets with both quantitative and categorical variables
- Simulate correlated variables using `mvrnorm()` from the MASS package
- Use `replicate()` and for loops to run simulations many times
- Evaluate how well simulated data recovers known population parameters
- Understand how sample size affects the precision of estimates

---

class: middle

# Building Simulated Datasets

---

## Combining Quantitative and Categorical Variables

We've seen how to generate random numbers and how to create categorical vectors with `rep()`. Now let's put them together.

```{r}
data.frame(
  group = rep(letters[1:2], each = 3),
  response = rnorm(n = 6, mean = 0, sd = 1)
)
```

--

.question[
How many observations does each group have? What distribution does `response` come from?
]

---

## Getting the Pattern Right Matters

When simulating **group differences**, the pattern of `rep()` must match how `rnorm()` cycles through its vector of means.

.pull-left[
### `rnorm()` cycles elementwise
```{r}
set.seed(42)
rnorm(n = 6, mean = c(5, 10), sd = 1)
```
The pattern: 5, 10, 5, 10, 5, 10
]

--

.pull-right[
### Match with `rep()`
```{r}
rep(letters[1:2], times = 3)
```

Not `each = 3`! That would give: a, a, a, b, b, b
]

---

## Building a Complete Dataset

```{r}
set.seed(42)
group <- rep(letters[1:2], times = 3)
response <- rnorm(n = 6, mean = c(5, 10), sd = 1)

data.frame(group, response)
```

--

.tip[
Building vectors separately before combining them into a data frame makes it easier to verify the patterns are correct.
]

---

## Adding Crossed Factors

What if we have two categorical variables that are **crossed** — every combination appears at least once?

```{r}
data.frame(
  group = rep(letters[1:2], each = 3),
  factor = rep(LETTERS[3:5], times = 2),
  response = rnorm(n = 6, mean = 0, sd = 1)
)
```

--

.question[
Why does `factor` use `times = 2` instead of `each = 2`?
]

---

## Watch Out for Recycling!

.pull-left[
### Lucky: An error
```{r error=TRUE}
data.frame(
  group = rep(LETTERS[3:4], each = 10),
  x = runif(n = 15, min = 10, max = 15)
)
```
R catches the mismatch!
]

--

.pull-right[
### Unlucky: Silent recycling
```{r}
df_bad <- data.frame(
  group = rep(LETTERS[3:4], each = 10),
  x = runif(n = 10, min = 10, max = 15)
)
```
R recycles the 10 values to fill 20 rows.
The first and last 10 rows are **identical**.
]

---

class: middle

# Welcome to Mars `r emo::ji("rocket")`

---

## The Mars Colony Simulation

.pull-left[
Imagine you are planning a Mars colony.

You need to simulate colonist attributes:
- **Age** — balancing youth with experience
- **Role** — engineer, scientist, medic
- **Health** — the MARSGAR score (0–100)
- **Skills** — technical, problem-solving
- **Personality** — Big Five traits
]

.pull-right[
<br>
```{r echo=FALSE, fig.align="center", out.width="85%"}
set.seed(123)
age <- rnorm(100, mean = 30, sd = 5)

data.frame(age) %>%
  ggplot(aes(x = age)) +
  geom_histogram(binwidth = 2, fill = viridis(5)[3], color = "black") +
  labs(title = "Age Distribution of 100 Colonists",
       x = "Age", y = "Count") +
  theme_minimal()
```
]

---

## Assigning Roles

There are many ways to assign categorical roles using `rep()` and `sample()`:

.midi[
```{r}
# Repeating pattern
roleA <- rep(c("engineer", "scientist", "medic"), length.out = 100)

# Blocks of each role
roleB <- rep(c("engineer", "scientist", "medic"), each = 34,
             length.out = 100)

# Random assignment with equal probability
set.seed(1)
roleD <- sample(c("engineer", "scientist", "medic"),
                size = 100, replace = TRUE)
```
]

--

.question[
Which method gives you exactly balanced groups? Which gives approximate balance?
]

---

## Comparing Assignment Methods

```{r echo=FALSE, fig.align="center", out.width="75%"}
set.seed(1)
roleA <- rep(c("engineer", "scientist", "medic"), length.out = 100)
roleB <- rep(c("engineer", "scientist", "medic"), each = 34, length.out = 100)
roleD <- sample(c("engineer", "scientist", "medic"), size = 100, replace = TRUE)

data.frame(
  role = c(roleA, roleB, roleD),
  method = rep(c("Repeating\n(length.out)", "Blocks\n(each)", "Random\n(sample)"),
               each = 100)
) %>%
  ggplot(aes(x = role, fill = method)) +
  geom_bar(position = position_dodge()) +
  scale_fill_viridis_d(end = 0.8) +
  labs(title = "Role Distribution by Assignment Method",
       x = "Role", y = "Count") +
  theme_minimal() +
  theme(legend.position = "bottom")
```

---

## Simulating Relationships Between Variables

So far, our variables have been **independent**. But real data has structure!

.pull-left[
### Independent (no relationship)
```{r eval=FALSE}
age <- rnorm(100, mean = 30, sd = 5)
health <- runif(100, min = 50, max = 100)
```
Age and health are unrelated by construction.
]

--

.pull-right[
### Dependent (built-in relationship)
```{r eval=FALSE}
age <- rnorm(100, mean = 30, sd = 5)
technical_skills <- 2 * age +
  rnorm(100, mean = 0, sd = 1)
```
Technical skills **depend on** age.
]

---

## Independent vs. Dependent: Visualized

```{r echo=FALSE, fig.align="center", out.width="85%"}
set.seed(123)
age <- rnorm(100, mean = 30, sd = 5)
health <- runif(100, min = 50, max = 100)
technical_skills <- 2 * age + rnorm(100, mean = 0, sd = 1)

bind_rows(
  data.frame(age, outcome = health, type = "Independent (Age vs MARSGAR)"),
  data.frame(age, outcome = technical_skills, type = "Dependent (Age vs Technical Skills)")
) %>%
  ggplot(aes(x = age, y = outcome)) +
  geom_point(color = viridis(3)[1], alpha = 0.7) +
  geom_smooth(method = "lm", se = FALSE, color = viridis(3)[2], formula = "y ~ x") +
  facet_wrap(~ type, scales = "free_y") +
  labs(x = "Age", y = "Outcome") +
  theme_minimal()
```

---

class: middle

# Simulating Correlated Data with `mvrnorm()`

---

## The Problem with One Variable at a Time

Building relationships with formulas like `y = 2*x + noise` works, but...

- You need to **know the equation** in advance
- Hard to control the **exact correlation**
- Doesn't scale well to many variables

--

.tip[
`mvrnorm()` from the **MASS** package lets you specify the correlation structure directly.
]

---

## How `mvrnorm()` Works

The function generates data from a **multivariate normal distribution**:

```{r eval=FALSE}
mvrnorm(n, mu, Sigma)
```

- `n` — number of observations
- `mu` — vector of means (one per variable)
- `Sigma` — covariance matrix (encodes variances and correlations)

--

.footnote[
Remember: a correlation matrix is a standardized covariance matrix where the diagonal is all 1s.
]

---

## A Simple Two-Variable Example

.pull-left[
```{r}
library(MASS)

mean_traits <- c(50, 50)
cov_matrix <- matrix(
  c(100, 50,
     50, 100),
  ncol = 2
)

set.seed(123)
traits <- mvrnorm(
  n = 100,
  mu = mean_traits,
  Sigma = cov_matrix
)
head(traits, 4)
```
]

.pull-right[
```{r echo=FALSE, fig.align="center", out.width="95%"}
colnames(traits) <- c("Resilience", "Agreeableness")
as.data.frame(traits) %>%
  ggplot(aes(x = Resilience, y = Agreeableness)) +
  geom_point(color = viridis(5)[2], alpha = 0.7) +
  geom_smooth(method = "lm", se = FALSE, color = viridis(5)[4], formula = "y ~ x") +
  labs(title = "Simulated Correlated Traits") +
  theme_minimal()
```
]

---

## From Correlation to Covariance

The correlation matrix is easier to think about, but `mvrnorm()` needs a **covariance** matrix.

.pull-left[
### Correlation matrix
$$r_{ij} = \frac{\text{Cov}(X_i, X_j)}{SD_i \cdot SD_j}$$

Diagonal is always 1.
]

.pull-right[
### Convert with R
```{r}
cor_mat <- matrix(c(1.0, 0.5,
                    0.5, 1.0), ncol = 2)
sd_vec <- c(10, 10)

# Correlation to Covariance
cov_mat <- cor_mat *
  (sd_vec %*% t(sd_vec))
cov_mat
```
]

---

## Checking Your Simulation

After generating data, check that it matches your parameters:

```{r}
set.seed(123)
sim_data <- mvrnorm(n = 100, mu = c(50, 50),
                    Sigma = cov_mat)
colnames(sim_data) <- c("V1", "V2")
```

.pull-left[
```{r}
# Check means
colMeans(sim_data)
```
```{r}
# Check SDs
apply(sim_data, 2, sd)
```
]
.pull-right[
```{r}
# Check correlation
cor(sim_data)
```
]

--

.footnote[
With n = 100, the sample statistics won't perfectly match the population parameters. That's sampling variability!
]

---

class: middle

# The Big Five on Mars

---

## Why Personality Matters for Mars

.pull-left[
The **Big Five** personality traits will shape colony life:

- **Extraversion (EX)** — social energy in isolation
- **Emotional Stability (ES)** — handling stress
- **Agreeableness (AG)** — resolving conflicts
- **Conscientiousness (CO)** — following protocols
- **Openness (OP)** — adapting to the unknown
]

.pull-right[
These traits are **correlated** in real populations. We can't simulate them independently!

```{r echo=FALSE, fig.align="center", out.width="85%"}
cor_matrix_bigfive <- matrix(
  c(1.00, 0.26, 0.20, 0.19, 0.29,
    0.26, 1.00, 0.16, 0.23, 0.07,
    0.20, 0.16, 1.00, 0.29, 0.20,
    0.19, 0.23, 0.29, 1.00, 0.16,
    0.29, 0.07, 0.20, 0.16, 1.00),
  nrow = 5, byrow = TRUE,
  dimnames = list(
    c("EX", "ES", "AG", "CO", "OP"),
    c("EX", "ES", "AG", "CO", "OP"))
)

as.data.frame(as.table(cor_matrix_bigfive)) %>%
  ggplot(aes(x = Var1, y = Var2, fill = Freq)) +
  geom_tile() +
  geom_text(aes(label = round(Freq, 2)), color = "white", size = 5) +
  scale_fill_viridis(option = "D") +
  labs(title = "Big Five Correlation Matrix",
       subtitle = "From Park et al. (2020)",
       x = "", y = "", fill = "r") +
  theme_minimal()
```
]

---

## Setting Up the Big Five Simulation

```{r}
# Population parameters
var_names <- c("EX", "ES", "AG", "CO", "OP")
mean_traits <- c(-0.5, 0.5, 0.25, 0.5, 0)
sd_traits <- c(1, 0.9, 1, 1, 1)

# Correlation matrix (Park et al., 2020)
cor_matrix_bigfive <- matrix(
  c(1.0000, 0.2599, 0.1972, 0.1860, 0.2949,
    0.2599, 1.0000, 0.1576, 0.2306, 0.0720,
    0.1972, 0.1576, 1.0000, 0.2866, 0.1951,
    0.1860, 0.2306, 0.2866, 1.0000, 0.1574,
    0.2949, 0.0720, 0.1951, 0.1574, 1.0000),
  nrow = 5, byrow = TRUE,
  dimnames = list(var_names, var_names)
)

# Convert to covariance matrix
cov_matrix_bigfive <- cor_matrix_bigfive * (sd_traits %*% t(sd_traits))
```

---

## Generating the Big Five Data

```{r}
set.seed(123)
sim_bigfive <- mvrnorm(
  n = 100,
  mu = mean_traits,
  Sigma = cov_matrix_bigfive
)
colnames(sim_bigfive) <- var_names
head(sim_bigfive, 5)
```

---

## How Well Did We Recover the Parameters?

.pull-left[
### Means
```{r echo=FALSE}
data.frame(
  Trait = var_names,
  Population = mean_traits,
  Sample = round(colMeans(sim_bigfive), 3)
) %>% kable()
```
]

.pull-right[
### SDs
```{r echo=FALSE}
data.frame(
  Trait = var_names,
  Population = sd_traits,
  Sample = round(apply(sim_bigfive, 2, sd), 3)
) %>% kable()
```
]

--

.footnote[
Close but not exact — that's sampling variability at work with n = 100.
]

---

## Recovered Correlations

```{r echo=FALSE, fig.align="center", out.width="65%"}
sim_cor <- cor(sim_bigfive)

# Create comparison data
pop_long <- as.data.frame(as.table(cor_matrix_bigfive)) %>%
  rename(Population = Freq)
sim_long <- as.data.frame(as.table(sim_cor)) %>%
  rename(Sample = Freq)

comparison <- left_join(pop_long, sim_long, by = c("Var1", "Var2")) %>%
  filter(as.numeric(Var1) < as.numeric(Var2))

ggplot(comparison, aes(x = Population, y = Sample)) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "gray50") +
  geom_point(size = 4, color = viridis(5)[2]) +
  geom_text(aes(label = paste0(Var1, "-", Var2)),
            vjust = -1, size = 3.5) +
  labs(title = "Population vs. Sample Correlations",
       subtitle = "Points on the dashed line = perfect recovery",
       x = "Population Correlation", y = "Sample Correlation") +
  coord_equal(xlim = c(0, 0.4), ylim = c(0, 0.4)) +
  theme_minimal()
```

---

class: middle

# Running Simulations Many Times

---

## Why Replicate?

A single simulation is like a single sample — it's subject to randomness.

To understand **long-run behavior**, we repeat the process many times.

--

Two approaches:
1. **`replicate()`** — concise, outputs a list or matrix
2. **`for()` loop** — explicit, more flexible

---

## Using `replicate()` for Many Colonies

```{r}
set.seed(124)
colonies <- replicate(
  n = 100,
  expr = mvrnorm(n = 100, mu = mean_traits,
                 Sigma = cov_matrix_bigfive),
  simplify = FALSE
)
```

--

Each element in `colonies` is a 100 x 5 matrix — one simulated planet's worth of colonists.

```{r}
length(colonies)
dim(colonies[[1]])
```

---

## Extracting Summary Statistics

Across 100 colonies, how does the **mean extraversion** vary?

```{r}
ex_means <- sapply(colonies, function(x) mean(x[, 1]))
```

.pull-left[
```{r}
summary(ex_means)
```
]
.pull-right[
```{r echo=FALSE, fig.align="center", out.width="95%"}
data.frame(ex_means) %>%
  ggplot(aes(x = ex_means)) +
  geom_histogram(binwidth = 0.05, fill = viridis(5)[3], color = "black") +
  geom_vline(xintercept = mean_traits[1], linetype = "dashed",
             color = viridis(5)[1], linewidth = 1.2) +
  annotate("text", x = mean_traits[1] + 0.15, y = Inf,
           label = paste0("Population\nmean = ", mean_traits[1]),
           vjust = 1.5, color = viridis(5)[1], size = 4) +
  labs(title = "Distribution of Mean EX Across 100 Colonies",
       x = "Mean Extraversion", y = "Count") +
  theme_minimal()
```
]

---

## Extracting Correlations Across Colonies

How stable is the EX–OP correlation across different colonies?

```{r}
ex_op_cors <- sapply(colonies, function(x) cor(x[, 1], x[, 5]))
```

```{r echo=FALSE, fig.align="center", out.width="60%"}
data.frame(r = ex_op_cors) %>%
  ggplot(aes(x = r)) +
  geom_histogram(binwidth = 0.03, fill = viridis(5)[4], color = "black") +
  geom_vline(xintercept = cor_matrix_bigfive[1, 5], linetype = "dashed",
             color = viridis(5)[1], linewidth = 1.2) +
  annotate("text", x = cor_matrix_bigfive[1, 5] + 0.08, y = Inf,
           label = paste0("Population\nr = ", round(cor_matrix_bigfive[1, 5], 3)),
           vjust = 1.5, color = viridis(5)[1], size = 4) +
  labs(title = "Distribution of EX-OP Correlation Across 100 Colonies",
       x = "Correlation (r)", y = "Count") +
  theme_minimal()
```

---

## The `for()` Loop Alternative

```{r}
set.seed(124)
all_sims <- data.frame()

for (i in 1:100) {
  sim_data <- mvrnorm(n = 100, mu = mean_traits,
                      Sigma = cov_matrix_bigfive)
  sim_df <- cbind.data.frame(
    colonist_id = 1:100, rep = i, sim_data
  )
  all_sims <- rbind(all_sims, sim_df)
}
```

```{r}
dim(all_sims)
```

--

.tip[
The `for()` loop is more explicit. Use whichever approach you find easier to read and debug.
]

---

class: middle

# How Much Data Do We Need?

---

## Sample Size and Estimation Precision

.pull-left[
With small samples, our estimates are **noisy**.

As sample size increases:
- Means converge to population values
- Correlations stabilize
- Covariances become more precise

Let's see this in action!
]

.pull-right[
```{r echo=FALSE, fig.align="center", out.width="95%"}
set.seed(124)
sample_sizes <- seq(30, 300, by = 15)
reps <- 20

results <- data.frame()
for (size in sample_sizes) {
  for (r in 1:reps) {
    d <- mvrnorm(n = size, mu = c(50, 50),
                 Sigma = matrix(c(100, 50, 50, 100), ncol = 2))
    results <- rbind(results,
      data.frame(SampleSize = size, Repetition = r,
                 Covariance = cov(d[, 1], d[, 2])))
  }
}

ggplot(results, aes(x = factor(SampleSize), y = Covariance)) +
  geom_boxplot(fill = viridis(3)[2], alpha = 0.7, outlier.size = 1) +
  geom_hline(yintercept = 50, linetype = "dashed", color = viridis(3)[1],
             linewidth = 1) +
  annotate("text", x = 1, y = 52, label = "Population = 50",
           hjust = 0, color = viridis(3)[1], size = 3.5) +
  labs(title = "Covariance Estimates by Sample Size",
       x = "Sample Size", y = "Estimated Covariance") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 8))
```
]

---

## The Distribution Narrows with More Data

```{r echo=FALSE, fig.align="center", out.width="75%"}
# Show just a few representative sizes
show_sizes <- c(30, 75, 150, 300)
results_subset <- results %>% filter(SampleSize %in% show_sizes)

ggplot(results_subset, aes(x = Covariance, fill = factor(SampleSize))) +
  geom_density(alpha = 0.6) +
  geom_vline(xintercept = 50, linetype = "dashed", linewidth = 1) +
  scale_fill_viridis_d(end = 0.85) +
  labs(title = "Covariance Estimates Converge as n Increases",
       x = "Estimated Covariance", y = "Density",
       fill = "Sample\nSize") +
  theme_minimal() +
  theme(legend.position = "bottom")
```

--

.footnote[
This is the **law of large numbers** in action: larger samples give estimates closer to the truth.
]

---

class: middle

# Simulation Workflow

---

## Putting It All Together

```
1. Define population parameters
   (means, SDs, correlations)
         |
2. Generate one dataset
   (rnorm, runif, mvrnorm)
         |
3. Check: does it match your parameters?
   (colMeans, cor, apply)
         |
4. Replicate many times
   (replicate or for loop)
         |
5. Analyze the distribution of results
   (histograms, summary stats)
```

---

class: middle

# Wrapping Up...

---

# Sources

- Ariel Muldoon's [tutorial](https://github.com/aosmith16/simulation-helper-functions)
- Park et al. (2020). Meta-analytic five-factor model personality intercorrelations. *Journal of Applied Psychology*, 105(12), 1490–1529.
- Mine Çetinkaya-Rundel's Data Science in a Box ([link](https://datasciencebox.org/))

---

class: middle

# Summary: Learning Goals Achieved

---

## What We've Learned

Today, you should now be able to...

.pull-left[
### Concepts
- Combining quantitative and categorical variables into datasets
- Simulating **correlated** variables with `mvrnorm()`
- Sampling variability: one simulation ≠ the truth
- More data → more precise estimates
]

.pull-right[
### Skills
- Building data frames with `rep()`, `rnorm()`, `runif()`
- Converting correlation matrices to covariance matrices
- Running repeated simulations with `replicate()` and `for()`
- Extracting and visualizing summary statistics across replications
]

---

class: middle

# Wrapping Up...

---
