---
title: "Simulating Correlated Data <br> `r emo::ji('game_die')`"
author: "S. Mason Garrison"
output:
  xaringan::moon_reader:
    css: "../slides.css"
    lib_dir: libs
    nature:
      ratio: "16:9"
      highlightLines: true
      highlightStyle: solarized-light
      countIncrementalSlides: false
      slideNumberFormat: ""
---



```{r child = "../setup.Rmd"}
```

```{r packages, echo = FALSE, message=FALSE, warning=FALSE}
library(tidyverse)
library(MASS)
library(emo)
library(viridis)
library(knitr)
# Remember to compile
# xaringan::inf_mr(cast_from = "..")
```

class: middle

# Learning Goals

---

## Learning Goals

By the end of this session, you will be able to...

- Explain why and when simulation is a useful tool in data science and psychology
- Simulate correlated variables using `mvrnorm()` from the MASS package
- Use simulation to demonstrate statistical concepts like the Central Limit Theorem
- Conduct a basic power analysis through simulation
- Evaluate how sample size affects the precision and reliability of estimates

---

class: middle

# Why Simulate?

---

## Last Time vs. Today

.pull-left[
Last time, we learned the **mechanics** of simulation:
- `rnorm()`, `runif()`, `rpois()`
- `rep()` for categorical variables
- `replicate()` and `for()` loops
]

--

.pull-right[
Today, we ask: **what can we actually do with all of this?**

Simulation lets us...
- Develop intuition about chance
- Demonstrate statistical principles
- Plan studies before collecting data
- Test whether our methods actually work
]

--

.hand[.light-blue[We have the tools. Now let's put them to work.]]

---

## The Big Picture

.alert[Have you ever found a "significant" result and wondered... is this real, or did I just get lucky?]

--

Simulation gives us a way to answer that question **directly**.

--

Instead of relying on formulas, we can:

1. Create a world where we **know** the truth

--

2. Generate data from that world

--

3. See how our methods perform

--

4. Repeat thousands of times

---

class: middle

# Spurious Correlations

---

## The Danger of Small Samples

.question[
You collect data on two variables from 10 people and find r = 0.45. Is that a real relationship?
]

--

Let's ask: **how often would we see r = 0.45 or more from two completely unrelated variables?**

---

## Simulating Spurious Correlations

.midi[
```{r}
set.seed(2024)
spurious_cors <- replicate(10000, {
  x <- rnorm(10)
  y <- rnorm(10)
  cor(x, y)
})
mean(abs(spurious_cors) >= 0.45)
```
]

--

About `r round(mean(abs(spurious_cors) >= 0.45) * 100)`% of the time!

--

.hand[.light-blue[With n = 10, even moderate correlations are expected by chance alone.]]

.footnote[
See [Spurious Correlations](https://www.tylervigen.com/spurious-correlations) for fun real-world examples.
]

---

## Spurious Correlations: Visualized

```{r echo=FALSE, fig.align="center", out.width="75%"}
data.frame(r = spurious_cors) %>%
  ggplot(aes(x = r)) +
  geom_histogram(binwidth = 0.05, fill = viridis(5)[3], color = "black") +
  geom_vline(xintercept = c(-0.45, 0.45), linetype = "dashed",
             color = viridis(5)[1], linewidth = 1) +
  annotate("text", x = 0.55, y = Inf, label = "|r| >= 0.45",
           vjust = 1.5, color = viridis(5)[1], size = 4) +
  labs(title = "Correlations Between Two Unrelated Variables (n = 10)",
       subtitle = "10,000 simulations from independent normals",
       x = "Correlation (r)", y = "Count") +
  theme_minimal()
```

---

## Let's Watch It Happen

Here's **one** random sample of n = 10 from two unrelated variables:

```{r echo=FALSE, fig.align="center", out.width="50%"}
set.seed(77)
d1 <- data.frame(x = rnorm(10), y = rnorm(10))
ggplot(d1, aes(x, y)) +
  geom_point(size = 3, color = viridis(5)[2]) +
  geom_smooth(method = "lm", se = FALSE, color = viridis(5)[4], formula = "y ~ x") +
  labs(title = paste0("Sample 1: r = ", round(cor(d1$x, d1$y), 2)),
       x = "X (random)", y = "Y (random)") +
  theme_minimal()
```

---

## Another Random Sample

Here's a **second** random sample — same population (no relationship):

```{r echo=FALSE, fig.align="center", out.width="50%"}
set.seed(88)
d2 <- data.frame(x = rnorm(10), y = rnorm(10))
ggplot(d2, aes(x, y)) +
  geom_point(size = 3, color = viridis(5)[3]) +
  geom_smooth(method = "lm", se = FALSE, color = viridis(5)[4], formula = "y ~ x") +
  labs(title = paste0("Sample 2: r = ", round(cor(d2$x, d2$y), 2)),
       x = "X (random)", y = "Y (random)") +
  theme_minimal()
```

--

Different sample, different correlation — and the true value is **zero** in both cases.

---

## And a Third...

```{r echo=FALSE, fig.align="center", out.width="50%"}
set.seed(99)
d3 <- data.frame(x = rnorm(10), y = rnorm(10))
ggplot(d3, aes(x, y)) +
  geom_point(size = 3, color = viridis(5)[1]) +
  geom_smooth(method = "lm", se = FALSE, color = viridis(5)[4], formula = "y ~ x") +
  labs(title = paste0("Sample 3: r = ", round(cor(d3$x, d3$y), 2)),
       x = "X (random)", y = "Y (random)") +
  theme_minimal()
```

--

.hand[.light-blue[We could keep going...]]

---

## 10,000 Samples Later

```{r echo=FALSE, fig.align="center", out.width="75%"}
data.frame(r = spurious_cors) %>%
  ggplot(aes(x = r)) +
  geom_histogram(binwidth = 0.05, fill = viridis(5)[3], color = "black") +
  labs(title = "Distribution of r Across 10,000 Samples (n = 10)",
       subtitle = "Population correlation = 0",
       x = "Correlation (r)", y = "Count") +
  theme_minimal()
```

Each bar represents thousands of samples — all from a world with **no relationship**.

---

## What About Larger Samples?

```{r echo=FALSE, fig.align="center", out.width="80%"}
set.seed(2024)
sim_cors <- data.frame()
for (n in c(10, 30, 100, 500)) {
  cors <- replicate(5000, cor(rnorm(n), rnorm(n)))
  sim_cors <- rbind(sim_cors, data.frame(n = n, r = cors))
}

ggplot(sim_cors, aes(x = r, fill = factor(n))) +
  geom_histogram(binwidth = 0.03, color = "black", alpha = 0.8) +
  facet_wrap(~ paste("n =", n), scales = "free_y") +
  scale_fill_viridis_d(end = 0.85) +
  guides(fill = "none") +
  labs(title = "Spurious Correlations Shrink with Sample Size",
       subtitle = "Correlations between two independent variables",
       x = "Correlation (r)", y = "Count") +
  theme_minimal()
```

--

.footnote[
This is why we need large samples — and why p-values exist. But simulation lets us **see** this behavior directly.
]

---

class: middle

# Wrapping Up...

---

class: middle

# The Central Limit Theorem via Simulation

---

## The Central Limit Theorem

.hand[
.light-blue[
The sampling distribution of the mean approaches a normal distribution as n increases, regardless of the population distribution.
]
]

--

That's a big claim! Let's test it with a decidedly non-normal distribution.

---

## Start with a Skewed Distribution

The Poisson distribution with a small lambda is **right-skewed**:

.pull-left[
```{r echo=FALSE, fig.align="center", out.width="95%"}
set.seed(123)
data.frame(x = rpois(1000, lambda = 2)) %>%
  ggplot(aes(x = x)) +
  geom_bar(fill = viridis(5)[4], color = "black") +
  labs(title = "Poisson Distribution (lambda = 2)",
       x = "Value", y = "Count") +
  theme_minimal()
```
]

.pull-right[
This is clearly **not** normal — it's discrete, bounded at 0, and skewed right.

--

.question[
What do you think happens when we take the **mean** of samples from this distribution?
]
]

---

## CLT in Action

```{r echo=FALSE, fig.align="center", out.width="80%"}
set.seed(42)
clt_data <- data.frame()
for (n in c(1, 5, 30, 100)) {
  means <- replicate(5000, mean(rpois(n, lambda = 2)))
  clt_data <- rbind(clt_data, data.frame(n = n, sample_mean = means))
}

ggplot(clt_data, aes(x = sample_mean, fill = factor(n))) +
  geom_histogram(binwidth = 0.1, color = "black", alpha = 0.8) +
  facet_wrap(~ paste("n =", n), scales = "free") +
  scale_fill_viridis_d(end = 0.85) +
  guides(fill = "none") +
  labs(title = "Sampling Distribution of the Mean (Poisson, lambda = 2)",
       subtitle = "5,000 replications at each sample size",
       x = "Sample Mean", y = "Count") +
  theme_minimal()
```

--

.footnote[
Even with n = 30, the distribution of means looks approximately normal — despite the population being skewed and discrete!
]

---

## The CLT Code

.midi[
```{r eval=FALSE}
set.seed(42)
sample_means <- replicate(
  n = 5000,
  expr = mean(rpois(n = 30, lambda = 2))
)

data.frame(sample_means) %>%
  ggplot(aes(x = sample_means)) +
  geom_histogram(binwidth = 0.1)
```
]

--

.question[
What would happen if you replaced `rpois(30, lambda = 2)` with `runif(30, min = 0, max = 1)`? Would the CLT still hold?
]

--

.tip[
Yes! Try it. The CLT holds for **any** population distribution, as long as n is large enough. That's what makes it so powerful.
]

---

class: middle

# Wrapping Up...

---

class: middle

# Simulating Correlated Data with `mvrnorm()`

---

## The Problem with One Variable at a Time

.alert[Imagine you're simulating data for a study on exam performance. You need midterm and final exam scores that are **correlated** — students who do well on the midterm tend to do well on the final.]

--

Building relationships with `y = 0.5*x + noise` works, but...

- You need to **know the equation** in advance
- Hard to control the **exact correlation**
- Doesn't scale well to 3, 4, 5+ variables

--

.tip[
`mvrnorm()` from the **MASS** package lets you specify the correlation structure directly.
]

---

## What Does `mvrnorm()` Need?

`mvrnorm()` takes three arguments:

```{r eval=FALSE}
mvrnorm(n, mu, Sigma)
```

--

- **`n`** — number of observations (e.g., 200 students)

--

- **`mu`** — vector of means (one per variable)

--

- **`Sigma`** — **covariance matrix** (variances on diagonal, covariances off-diagonal)

--

.footnote[
A correlation matrix is just a standardized covariance matrix where the diagonal is all 1s.
]

---

## Step 1: Define the Means

Let's simulate midterm and final exam scores.

Both exams have a mean of 100:

```{r}
mu <- c(100, 100)
```

--

.midi[
Think of this as: "On average, students score 100 on both exams."
]

---

## Step 2: Build the Covariance Matrix

We want a correlation of r = 0.4 between the exams, with SD = 15 for both.

--

First, start with the **correlation matrix**:

```{r}
cor_mat <- matrix(c(1.0, 0.4,
                    0.4, 1.0), ncol = 2)
cor_mat
```

---

## Step 2 (cont'd): Convert to Covariance

The formula: $\text{Cov}(X_i, X_j) = r_{ij} \cdot SD_i \cdot SD_j$

--

```{r}
sd_vec <- c(15, 15)

# Correlation -> Covariance
cov_mat <- cor_mat * (sd_vec %*% t(sd_vec))
cov_mat
```

--

.midi[
The diagonal values (225) are the **variances** ($15^2 = 225$). The off-diagonal values (90) are the **covariances** ($0.4 \times 15 \times 15 = 90$).
]

---

## Step 3: Generate the Data

```{r}
library(MASS)

set.seed(11)
sim <- mvrnorm(n = 200, mu = mu, Sigma = cov_mat)
head(sim, 5)
```

--

We have 200 simulated students with two exam scores each!

---

## Step 4: Plot and Interpret

```{r echo=FALSE, fig.align="center", out.width="65%"}
colnames(sim) <- c("Midterm", "Final")
as.data.frame(sim) %>%
  ggplot(aes(x = Midterm, y = Final)) +
  geom_point(color = viridis(5)[2], alpha = 0.6) +
  geom_smooth(method = "lm", se = FALSE, color = viridis(5)[4], formula = "y ~ x") +
  labs(title = "Simulated Exam Scores (r = 0.4)",
       x = "Midterm Score", y = "Final Score") +
  theme_minimal()
```

--

Students who scored higher on the midterm tend to score higher on the final — just as we specified.

---

## Checking Your Work

.question[
How do we know the simulated data actually matches what we asked for?
]

--

Always verify your parameters after generating data:

.pull-left[
```{r}
# Check means (target: 100, 100)
colMeans(sim)
```
```{r}
# Check SDs (target: 15, 15)
apply(sim, 2, sd)
```
]
.pull-right[
```{r}
# Check correlation (target: 0.4)
cor(sim)
```
]

--

.footnote[
Close but not exact — that's sampling variability! With n = 200, we get *estimates*, not the exact population values.
]

---

## Scaling Up: Four Exam Scores

`mvrnorm()` scales naturally. Just build a bigger matrix.

.small[
```{r}
var_names <- c("Quiz", "Midterm", "Project", "Final")
mu <- c(75, 70, 80, 72)
sd_exams <- c(10, 12, 8, 15)

cor_exams <- matrix(
  c(1.00, 0.55, 0.35, 0.50,
    0.55, 1.00, 0.40, 0.65,
    0.35, 0.40, 1.00, 0.45,
    0.50, 0.65, 0.45, 1.00),
  nrow = 4, byrow = TRUE,
  dimnames = list(var_names, var_names)
)

cov_exams <- cor_exams * (sd_exams %*% t(sd_exams))
```
]

--

.tip[
When building correlation matrices, make sure they are **symmetric** (same value above and below the diagonal) and **positive definite** (all eigenvalues > 0). R will throw an error if the matrix isn't valid.
]

---

## Did It Work?

```{r echo=FALSE, fig.align="center", out.width="65%"}
set.seed(42)
exam_data <- mvrnorm(n = 150, mu = mu, Sigma = cov_exams)
colnames(exam_data) <- var_names

sim_cor <- cor(exam_data)
pop_long <- as.data.frame(as.table(cor_exams)) %>% rename(Population = Freq)
sim_long <- as.data.frame(as.table(sim_cor)) %>% rename(Sample = Freq)

comparison <- left_join(pop_long, sim_long, by = c("Var1", "Var2")) %>%
  filter(as.numeric(Var1) < as.numeric(Var2))

ggplot(comparison, aes(x = Population, y = Sample)) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "gray50") +
  geom_point(size = 4, color = viridis(5)[2]) +
  geom_text(aes(label = paste0(Var1, "-", Var2)),
            vjust = -1, size = 3.5) +
  labs(title = "Population vs. Sample Correlations (n = 150)",
       subtitle = "Points on the dashed line = perfect recovery",
       x = "Population Correlation", y = "Sample Correlation") +
  coord_equal(xlim = c(0.2, 0.8), ylim = c(0.2, 0.8)) +
  theme_minimal()
```

--

The recovered correlations are close to the population values — simulation is working as intended.

---

class: middle

# Wrapping Up...

---

class: middle

# Power Analysis via Simulation

---

## What is Statistical Power?

**Power** = the probability of detecting an effect *when one truly exists*.

--

- Low power means we miss real effects (waste of time and resources)
- High power means we reliably detect effects (good science)

--

.question[
You're planning a study comparing reaction times between two groups. How many participants do you need?
]

--

We can answer this **entirely through simulation** — no formulas needed.

---

## Building a Power Simulation, Step by Step

First, let's think about what a single simulated experiment looks like:

```{r eval=FALSE}
# One simulated experiment
group1 <- rnorm(n_per_group, mean = 0, sd = 10)
group2 <- rnorm(n_per_group, mean = 5, sd = 10)
t.test(group1, group2)$p.value
```

--

We generate two groups with a **known** difference (5 points), then test whether we can detect it.

---

## Repeating the Experiment

Now repeat that experiment many times:

```{r eval=FALSE}
results <- replicate(1000, {
  group1 <- rnorm(n_per_group, mean = 0, sd = 10)
  group2 <- rnorm(n_per_group, mean = 5, sd = 10)
  t.test(group1, group2)$p.value < 0.05
})
```

--

Each replication returns `TRUE` (significant) or `FALSE` (not significant).

---

## Wrapping It in a Function

```{r}
power_sim <- function(n_per_group, effect = 5, sd = 10,
                      alpha = 0.05, nsim = 1000) {
  significant <- replicate(nsim, {
    group1 <- rnorm(n_per_group, mean = 0, sd = sd)
    group2 <- rnorm(n_per_group, mean = effect, sd = sd)
    t.test(group1, group2)$p.value < alpha
  })
  mean(significant)
}
```

--

The output is the proportion of significant results — that's **power**.

---

## Let's Try a Few Sample Sizes

```{r}
set.seed(2024)
power_sim(n_per_group = 10)
```

--

```{r}
power_sim(n_per_group = 30)
```

--

```{r}
power_sim(n_per_group = 60)
```

--

.midi[
Power goes up as sample size increases. But where does it cross the conventional 80% threshold?
]

---

## The Power Curve

```{r echo=FALSE, fig.align="center", out.width="80%"}
set.seed(2024)
sample_sizes <- seq(10, 100, by = 5)
power_results <- sapply(sample_sizes, power_sim)

data.frame(n = sample_sizes, power = power_results) %>%
  ggplot(aes(x = n, y = power)) +
  geom_line(color = viridis(3)[2], linewidth = 1.2) +
  geom_point(color = viridis(3)[2], size = 2) +
  geom_hline(yintercept = 0.80, linetype = "dashed", color = viridis(3)[1]) +
  annotate("text", x = 15, y = 0.82, label = "80% power",
           color = viridis(3)[1], size = 4, hjust = 0) +
  labs(title = "Statistical Power by Sample Size",
       subtitle = "Two-group comparison, d = 0.5 (effect = 5, SD = 10)",
       x = "n per group", y = "Power") +
  scale_y_continuous(labels = scales::percent, limits = c(0, 1)) +
  theme_minimal()
```

--

.footnote[
We need about `r sample_sizes[which.min(abs(power_results - 0.80))]` participants per group for 80% power. No formula required — just simulation!
]

---

## Why Simulation-Based Power Analysis?

.pull-left[
### Formulas work for simple cases
- Two groups, equal variances
- One predictor, normal outcome
- Standard tests (t, ANOVA, chi-square)
]

--

.pull-right[
### Simulation works for everything
- Complex designs (repeated measures, nested data)
- Non-normal data (e.g., reaction time distributions)
- Missing data patterns
- Novel statistical methods
- Any model you can fit in R
]

--

.footnote[
This approach is standard in quantitative psychology. See Fan (2012) in the *APA Handbook of Research Methods* for a great introduction to designing simulation studies.
]

---

class: middle

# Wrapping Up...

---

class: middle

# How Much Data Do We Need?

---

## Sample Size and Precision

.question[
If we increase our sample size, do we get more precise estimates? How much more precise?
]

--

Let's go back to our exam score simulation (r = 0.4) and see how the estimate varies across sample sizes.

---

## Watching Estimates Converge

```{r echo=FALSE, fig.align="center", out.width="80%"}
set.seed(124)
sample_sizes <- seq(30, 300, by = 30)
reps <- 50

results <- data.frame()
for (size in sample_sizes) {
  for (r in 1:reps) {
    d <- mvrnorm(n = size, mu = c(100, 100),
                 Sigma = matrix(c(225, 90, 90, 225), ncol = 2))
    results <- rbind(results,
      data.frame(SampleSize = size, Repetition = r,
                 Correlation = cor(d[, 1], d[, 2])))
  }
}

ggplot(results, aes(x = factor(SampleSize), y = Correlation)) +
  geom_boxplot(fill = viridis(3)[2], alpha = 0.7, outlier.size = 1) +
  geom_hline(yintercept = 0.4, linetype = "dashed", color = viridis(3)[1],
             linewidth = 1) +
  annotate("text", x = 1, y = 0.42, label = "Population r = 0.4",
           hjust = 0, color = viridis(3)[1], size = 3.5) +
  labs(title = "Correlation Estimates by Sample Size",
       x = "Sample Size", y = "Estimated Correlation") +
  theme_minimal()
```

---

## The Distribution Narrows

```{r echo=FALSE, fig.align="center", out.width="75%"}
show_sizes <- c(30, 90, 150, 300)
results_subset <- results %>% filter(SampleSize %in% show_sizes)

ggplot(results_subset, aes(x = Correlation, fill = factor(SampleSize))) +
  geom_density(alpha = 0.6) +
  geom_vline(xintercept = 0.4, linetype = "dashed", linewidth = 1) +
  scale_fill_viridis_d(end = 0.85) +
  labs(title = "Correlation Estimates Converge as n Increases",
       x = "Estimated Correlation", y = "Density",
       fill = "Sample\nSize") +
  theme_minimal() +
  theme(legend.position = "bottom")
```

--

.hand[.light-blue[Larger samples give estimates closer to the population value — the law of large numbers in action.]]

---

class: middle

# Wrapping Up...

---

class: middle

# Monte Carlo Simulation as a Research Tool

---

## What is a Monte Carlo Study?

.alert[You've developed a new statistical method. How do you know it actually works?]

--

You can't test it on real data — because with real data, you never know the **truth**.

--

A **Monte Carlo simulation study** is a computational experiment where you:

--

1. Define a **data-generating process** with known properties

--

2. Generate many datasets from that process

--

3. Apply your statistical method to each dataset

--

4. Evaluate how well the method performs

--

.hand[
.light-blue[
It's like a laboratory experiment, but for statistical methods.
]
]

---

## The Simulation Workflow

.question[
What are the steps for designing a Monte Carlo study?
]

--

**Step 1:** Define population parameters (means, SDs, correlations, effects)

--

**Step 2:** Generate one dataset (`rnorm`, `runif`, `mvrnorm`)

--

**Step 3:** Apply your method and extract results (`lm`, `t.test`, `cor`, ...)

--

**Step 4:** Repeat many times (`replicate` or `for` loop)

--

**Step 5:** Summarize across replications (bias, power, coverage, etc.)

--

.tip[
By saving results into a list, you can use `lapply()` or `purrr::map()` to extract whatever summaries you need — means, correlations, p-values, confidence intervals, anything.
]

---

## Use Cases in Psychology

Monte Carlo simulation is used **all the time** in quantitative psychology:

--

- Does my method give unbiased estimates? Compare sample averages to known population values

--

- How often does my test reject when it should? **Power**

--

- How often does my test reject when it shouldn't? **Type I error rate**

--

- What happens when assumptions are violated? **Robustness**

--

- How large a sample do I need? **Sample size planning**

--

.footnote[
For a deeper dive, see Muthén & Muthén (2002) on using Monte Carlo simulation for sample size and power determination, and Fan (2012) in the APA Handbook for a wonderful introduction aimed at psychology researchers.
]

---

class: middle

# Wrapping Up...

---

# Sources

- Ariel Muldoon's [tutorial](https://github.com/aosmith16/simulation-helper-functions)
- Mine Çetinkaya-Rundel's Data Science in a Box ([link](https://datasciencebox.org/))
- Fan, X. (2012). Designing simulation studies. In H. Cooper (Ed.), *APA Handbook of Research Methods in Psychology* (Vol. 2, pp. 427-444). APA.

---

class: middle

# Summary: Learning Goals Achieved

---

## What We've Learned

Today, you should now be able to...

.pull-left[
### Concepts
- ✅ Simulation as a way of thinking
- ✅ Spurious correlations and sample size
- ✅ Central Limit Theorem
- ✅ Power analysis
- ✅ Monte Carlo methodology
]

.pull-right[
### Skills
- ✅ Simulate correlated data with `mvrnorm()`
- ✅ Convert correlation to covariance matrices
- ✅ Run power analyses by simulation
- ✅ Design and evaluate simulation studies
]

---

class: middle

# Wrapping Up...

---
