---
title: "Getting Started <br> `r emo::ji('compass')`<br> Your Course Ecosystem"
author: "S. Mason Garrison"
output:
  xaringan::moon_reader:
    css: "../slides.css"
    lib_dir: libs
    self_contained: TRUE
    nature:
      ratio: "16:9"
      highlightLines: true
      highlightStyle: solarized-light
      countIncrementalSlides: false
---



```{r child = "../setup.Rmd"}
```

```{r packages, echo=FALSE, message=FALSE, warning=FALSE}
knitr::opts_chunk$set(knitr.duplicate.label = "allow")
library(tidyverse)
if (!require("emo")) devtools::install_github("hadley/emo")
library(emo)
```


class: middle

# Welcome to Your Extended Syllabus Day

.large[
Today we're going to set everything up together, step by step.
]

---

## Learning Goals

By the end of this session, you will be able to...

- Identify where all course materials live (spoiler: it's all GitHub `r emo::ji("winking_face")`)
- Navigate the course notes, syllabus, and slides
- Set up your GitHub account
- Install R and RStudio on your laptop
- Understand the flipped classroom workflow
- Know what a typical week looks like

---

class: middle

# "Wait, is everything on a different website?"

---

## The Big Reveal

.large[
Everything in this course lives on **GitHub**.
]

--

.pull-left[
What it **looks** like:
- "The syllabus website"
- "The course notes site"
- "Some GitHub thing for assignments"
- "The slides"
]

--

.pull-right[
What it **actually** is:
- GitHub Pages `r emo::ji("right_arrow")` smasongarrison.github.io/syllabi
- GitHub Pages `r emo::ji("right_arrow")` DataScience4Psych.github.io
- GitHub repos `r emo::ji("right_arrow")` github.com/DataScience4Psych
- GitHub Pages `r emo::ji("right_arrow")` (you're looking at them right now)
]

--

.center[.large[
`r emo::ji("bulb")` **One platform. Different views of the same place.**
]]

---

## How GitHub Pages Works

.center[
```
GitHub Repository (where files live)
        |
        v
GitHub Pages (turns files into a website)
        |
        v
You see a nice website at *.github.io
```
]

--

- The syllabus? Files in a GitHub repo, rendered as a website.
- The course notes? Files in a GitHub repo, rendered as a website.
- The slides? Files in a GitHub repo, rendered as a website.
- Your assignments? Files in a GitHub repo that **you** will work on.

--

.center[.large[
It's GitHub all the way down. `r emo::ji("turtle")`
]]
---

class: middle

# Let's Tour the Ecosystem

---

## `r emo::ji("one")` The Course Notes (Your Home Base)

.large[
.center[
[DataScience4Psych.github.io/DataScience4Psych/](https://DataScience4Psych.github.io/DataScience4Psych/)
]
]

--

- This is where the **content** lives: readings, tutorials, videos
- Organized by module (we'll walk through these in order)
- Includes links to labs, assignments, and the portfolio
- **Bookmark this.** You'll visit it every week.

--

.tip[
Think of the course notes as your **textbook replacement** -- but interactive, with embedded code and videos.
]

---

## `r emo::ji("two")` The Syllabus

.large[
.center[
[smasongarrison.github.io/syllabi/applieddatascience.html](https://smasongarrison.github.io/syllabi/applieddatascience.html)
]
]

--

- Policies, grading, schedule, office hours
- Contract grading details
- The "official" course information

--

.tip[
You'll reference the syllabus most at the **start** of the semester and when you have questions about grading or policies. The course notes are your day-to-day resource.
]

---

## `r emo::ji("three")` The GitHub Organization

.large[
.center[
[github.com/DataScience4Psych](https://github.com/DataScience4Psych)
]
]

--

- This is the GitHub **organization** for our class
- Contains repositories for:
  - Labs and assignments (e.g., `ae01a_unvotes`)
  - The course notes themselves
  - Anything else we build together

--

.tip[
When you get an assignment link, it will point here. You'll fork or clone the repo, do your work, and push it back.
]

---

## How the Pieces Connect

```
┌──────────────────────────────────────────────────┐
│              github.com/DataScience4Psych         │
│              (GitHub Organization)                │
│                                                   │
│  ┌──────────────┐  ┌──────────────┐  ┌────────┐ │
│  │ Course Notes  │  │  Labs & HW   │  │ Slides │ │
│  │    Repo       │  │   Repos      │  │  Repo  │ │
│  └──────┬───────┘  └──────────────┘  └────┬───┘ │
│         │                                  │      │
└─────────┼──────────────────────────────────┼──────┘
          │                                  │
          v                                  v
   Course Notes Website              Slide Decks
   (your home base)               (in-class content)
```

.center[
Plus: **smasongarrison.github.io/syllabi** `r emo::ji("right_arrow")` Syllabus (policies & grading)
]

---

class: middle

# Now Let's Get You Set Up

---

## Setup Checklist

We're going to do all of these **together, right now**:

.large[
1. `r emo::ji("check_box")` Create a GitHub account
2. `r emo::ji("check_box")` Install R
3. `r emo::ji("check_box")` Install RStudio
4. `r emo::ji("check_box")` Verify everything works
]

--

.center[.large[
`r emo::ji("laptop")` Open your laptops!
]]

---

class: middle

# Step 1: Create a GitHub Account

---

## Create a GitHub account

.instructions[
Go to [github.com](https://github.com/), and create an account (unless you already have one).
]

Tips for selecting a username:

.small[
- Incorporate your actual name.
- Reuse username from other contexts, e.g., Twitter or Slack.
- Pick a username you'll be comfortable revealing to your future boss.
- Shorter is better than longer.
- Be as unique as possible in as few characters as possible.
- Make it timeless. Don't highlight your current university, employer, etc.
- Avoid words laden with special meaning in programming, like `NA`.
]

.footnote[
Source: [Happy git with R](http://happygitwithr.com/github-acct.html#username-advice) by Jenny Bryan
]

---

## Once You Have an Account

.your-turn[
1. Go to [github.com/DataScience4Psych](https://github.com/DataScience4Psych)
2. Look around -- can you find the course notes repository?
3. Click on it and browse the files

You're looking at the **source code** behind the course notes website!
]

--

.question[
What do you notice? What kinds of files do you see?
]

---

class: middle

# Step 2: Install R

---

## Installing R

.instructions[
Go to [cloud.r-project.org](https://cloud.r-project.org/) and download R for your operating system.
]

.pull-left[
### Windows
- Click "Download R for Windows"
- Click "base"
- Click the download link
- Run the installer (defaults are fine)
]

.pull-right[
### macOS
- Click "Download R for macOS"
- Choose the version for your Mac (Apple Silicon or Intel)
- Download and install the `.pkg` file
]

--

.question[
Raise your hand when R is installed! `r emo::ji("raising_hand")`
]

---

class: middle

# Step 3: Install RStudio

---

## Installing RStudio

.instructions[
Go to [posit.co/download/rstudio-desktop/](https://posit.co/download/rstudio-desktop/) and download the free version.
]

--

- R is the **engine** `r emo::ji("engine")`
- RStudio is the **dashboard** `r emo::ji("car")`

You need both. You'll almost always open **RStudio**, not R directly.

--

.question[
Raise your hand when RStudio is installed! `r emo::ji("raising_hand")`
]

---

## Quick RStudio Tour

When you open RStudio, you'll see four panels:

.pull-left[
**Top Left:** Source/Editor
- Where you write code and documents

**Bottom Left:** Console
- Where code runs
- Try typing: `1 + 1`
]

.pull-right[
**Top Right:** Environment
- Shows your data and variables

**Bottom Right:** Files / Plots / Help
- File browser, plot viewer, help docs
]

---

## Let's Verify Everything Works

.your-turn[
Open RStudio and type the following in the **Console** (bottom left):

```r
print("Hello, Data Science!")
```

Then try:

```r
R.version.string
```
]

--

.question[
What version of R are you running? Type it in the chat!
]

---

class: middle

# The Flipped Classroom

---

## How This Class Actually Works

This is a **flipped classroom**. Here's what that means:

--

.pull-left[
### Before Class
- Watch pre-recorded lecture videos
- Read the relevant course notes module
- Work through the embedded examples
- Come with questions!
]

--

.pull-right[
### During Class
- **Tuesdays:** Face-to-face tutorials
  - Work through problems together
  - Hands-on coding
  - Get help in real time
- **Thursdays:** Solidarity sessions
  - Extended work time
  - Peer collaboration
]

---

## A Typical Week

```
Monday         Tuesday        Wednesday      Thursday       Friday
  │               │               │              │             │
  │  ┌────────┐   │               │  ┌────────┐  │             │
  │  │ Watch  │   │  In-Class     │  │Continue│  │ Solidarity  │
  │  │ Videos │   │  Tutorial     │  │ Work   │  │  Session    │
  │  │ + Read │   │  (hands-on)   │  │        │  │ (collab)    │
  │  │ Notes  │   │               │  │        │  │             │
  │  └────────┘   │               │  └────────┘  │             │
  │               │               │              │             │
  ▼               ▼               ▼              ▼             ▼
```

--

.tip[
The pre-recorded content **is** the lecture. Class time is for **doing**, not watching.
]

---

## Where to Find What (Quick Reference)

| I need to...                        | Go to...                          |
|-------------------------------------|-----------------------------------|
| Watch lectures / read content       | Course Notes                      |
| Check policies or grading           | Syllabus                          |
| Start a lab or assignment           | GitHub Organization               |
| Ask a question                      | Course Discussion Forum           |
| Get help from the professor         | Office Hours (Calendly)           |
| See the module schedule             | Course Notes or Syllabus          |

--

.center[.large[
When in doubt, start at the **Course Notes**. Everything links from there.
]]

---

class: middle

# Getting Help

---

## How to Get Help

.pull-left[
### Course Discussion Forum
- Post questions here (**not** DMs)
- Search before posting
- Others benefit from seeing Q&A
- Use proper code formatting!
]

.pull-right[
### Office Hours
- Remote via Zoom
- Book at [calendly.com/smasongarrison/](http://www.calendly.com/smasongarrison/)
- Come with specific questions
- No question is too small
]

---

## Tips for Asking Good Questions

.pull-left[
### Do `r emo::ji("check_mark")`
- "I'm getting this error when I try to filter: `Error: object 'x' not found`"
- Share the code that caused the issue
- Describe what you expected vs. what happened
]

.pull-right[
### Don't `r emo::ji("cross_mark")`
- "It doesn't work"
- "Help me with HW 2 Q3"
- Screenshots of code (copy-paste instead!)
]

--

.tip[
Give context from **course concepts**, not assignment numbers. "I have a question about filtering" is better than "I have a question about HW 1 Q4."
]

---

class: middle

# Contract Grading

---

## How Grading Works

- This course uses **contract grading**
- Assessment based on **effort and completion**, not perfection
- More representative of the real scientific process

--

### Your Milestones
.pull-left[
.large[
- Labs
- Portfolio
]
]

.pull-right[
Specifics are in the syllabus and course notes -- we'll revisit grading details as the semester goes on.
]

---

class: middle

# Let's Navigate Together

---

## Live Walk-Through

.your-turn[
Follow along on your laptop:

1. Open the **Course Notes**: [DataScience4Psych.github.io/DataScience4Psych/](https://DataScience4Psych.github.io/DataScience4Psych/)
2. Find the table of contents -- click through a few modules
3. Open the **Syllabus**: [smasongarrison.github.io/syllabi/applieddatascience.html](https://smasongarrison.github.io/syllabi/applieddatascience.html)
4. Find the grading section
5. Open the **GitHub Org**: [github.com/DataScience4Psych](https://github.com/DataScience4Psych)
6. Find a lab repository (e.g., `ae01a_unvotes`)
]

--

.question[
Can you see how these three places connect? What questions do you have?
]

---

## Quick Recap: Your Bookmarks

.large[
Save these three links:

1. `r emo::ji("book")` **Course Notes** (daily resource)
   - [DataScience4Psych.github.io/DataScience4Psych/](https://DataScience4Psych.github.io/DataScience4Psych/)

2. `r emo::ji("scroll")` **Syllabus** (policies & grading)
   - [smasongarrison.github.io/syllabi/applieddatascience.html](https://smasongarrison.github.io/syllabi/applieddatascience.html)

3. `r emo::ji("file_folder")` **GitHub Org** (assignments & source)
   - [github.com/DataScience4Psych](https://github.com/DataScience4Psych)
]

---

class: middle

# Summary: Learning Goals Achieved

---

## What We've Accomplished Today

.pull-left[
### Understanding
- `r emo::ji("check_mark")` All course materials live on GitHub
- `r emo::ji("check_mark")` Course notes = daily home base
- `r emo::ji("check_mark")` Syllabus = policies & grading
- `r emo::ji("check_mark")` GitHub Org = assignments & source
]

.pull-right[
### Setup Complete
- `r emo::ji("check_mark")` GitHub account created
- `r emo::ji("check_mark")` R installed
- `r emo::ji("check_mark")` RStudio installed
- `r emo::ji("check_mark")` Verified everything works
]

---

## What's Coming Next

- Next module: **What is Data Science?** (the fun stuff!)
- You'll watch pre-recorded content before our next class
- Start at the Course Notes, Module 1

--

.center[.large[
`r emo::ji("compass")` You now know where everything is.

`r emo::ji("laptop")` You have the tools installed.

`r emo::ji("rocket")` You're ready to go!
]]

---

class: middle

.hand[
Now that you know where everything lives...
<br>
let's do some data science!
]
