---
title: "Class Session Slides <br> `r emo::ji('cat')`"
author: "S. Mason Garrison"
output:
  xaringan::moon_reader:
    css: "../slides.css"
    lib_dir: libs
    self_contained: TRUE
    nature:
      ratio: "16:9"
      highlightLines: true
      highlightStyle: solarized-light
      countIncrementalSlides: false
---



```{r child = "../setup.Rmd"}
```


```{r packages, echo=FALSE, message=FALSE, warning=FALSE}
# Remember to compile
#xaringan::inf_mr(cast_from = "..")
#       slideNumberFormat: ""  
knitr::opts_chunk$set(knitr.duplicate.label = "allow")
library(tidyverse)
if (!require("emo")) devtools::install_github("hadley/emo")
library(emo)
if (!require("jasmines")) devtools::install_github("djnavarro/jasmines")
if (!require("mathart")) devtools::install_github("marcusvolz/mathart")
```


## Navigation Menu


### **Modules Overview**
- [Effective Data Visualization](#3)


---

## **Module 5: Effective Data Visualization**

### **Today’s Plan:**
- **Refine & Troubleshoot**: Improve visualizations based on feedback.
- **Analyze Misleading Visuals**: Discuss & revise problematic designs.
- **Spatial Data Visualization**: Hands-on lab (for those at that stage).


**By the end of this module, you should be able to:**
- Apply effective `ggplot2` techniques, including appropriate geoms, scales, and coordinate systems.
- Analyze misleading visualizations and discuss better design choices.
- Use spatial data visualization techniques in the lab.


[Back to Menu](#2)

