---
title: "Class Session Slides <br> `r emo::ji('cat')`"
author: "S. Mason Garrison"
output:
  xaringan::moon_reader:
    css: "../slides.css"
    lib_dir: libs
    self_contained: TRUE
    nature:
      ratio: "16:9"
      highlightLines: true
      highlightStyle: solarized-light
      countIncrementalSlides: false
---



```{r child = "../setup.Rmd"}
```


```{r packages, echo=FALSE, message=FALSE, warning=FALSE}
# Remember to compile
#xaringan::inf_mr(cast_from = "..")
#       slideNumberFormat: ""  
knitr::opts_chunk$set(knitr.duplicate.label = "allow")
library(tidyverse)
if (!require("emo")) devtools::install_github("hadley/emo")
library(emo)
if (!require("jasmines")) devtools::install_github("djnavarro/jasmines")
if (!require("mathart")) devtools::install_github("marcusvolz/mathart")
```


## Navigation Menu

### **Modules Overview**
- [M1: Welcome to Data Science](#3)
- [M2: Data and Visualization](#4)
- [M3: Welcome to the Tidyverse](#5)
- [M4: Data Diving with Types](#6)
- [M5: Effective Data Visualization](#7)
- [Check-In: Midterm Review](#8)
- [M6: Confounding and Communication](#9)
- [M7: Web Scraping](#10)

---

## **Module 1: Welcome to Data Science**

### **Today’s Plan:**
- **Introduction to Data Science**: Key concepts and importance.
- **Tools & Setup**: Overview of R, RStudio, and Git.
- **Exploring Data**: Initial data exploration techniques.
- **Hello R**: Hands-on lab (for those at that stage).

**By the end of this module, you should be able to:**
- Explain what data science is and why it matters.
- Set up and use basic tools for data science.
- Perform basic data exploration using R.

[Back to Menu](#2)

---

## **Module 2: Data and Visualization**

### **Today’s Plan:**
- **Exploratory Data Analysis (EDA)**: Understanding data distribution.
- **Principles of Visualization**: Designing clear and effective graphs.
- **Using `ggplot2`**: Building visualizations step-by-step.
- **Plastic Waste**: Hands-on lab (for those at that stage).

**By the end of this module, you should be able to:**
- Conduct exploratory data analysis in R.
- Design effective and clear data visualizations.
- Use `ggplot2` for creating different types of plots.

[Back to Menu](#2)
  
---

## **Module 3: Welcome to the Tidyverse**

### **Today’s Plan:**
- **Tidy Data**: Understanding structured data principles.
- **Grammar of Data Wrangling**: Key transformations with `dplyr`.
- **Working with Multiple Data Frames**: Efficient data merging techniques.
- **Nobel Laureates**: Hands-on lab (for those at that stage).

**By the end of this module, you should be able to:**
- Clean and reshape data using `tidyverse` tools.
- Use `dplyr` for common data wrangling tasks.
- Work with multiple datasets efficiently in R.

[Back to Menu](#2)

---

## **Module 4: Data Diving with Types**

### **Today’s Plan:**
- **Data Types & Recoding**: Understanding data structures.
- **Importing Data**: Reading and processing datasets in R.
- **Transformations & Cleaning**: Handling missing data and type conversions.
- **Visualizing Spatial Data**: Hands-on lab (for those at that stage).

**By the end of this module, you should be able to:**
- Understand different data types and their implications.
- Import and process data from various sources.
- Perform essential data transformations in R.

[Back to Menu](#2)
---
  
## **Module 5: Effective Data Visualization**

### **Today’s Plan:**
- **Refine & Troubleshoot**: Improve visualizations based on feedback.
- **Analyze Misleading Visuals**: Discuss & revise problematic designs.
- **Spatial Data Visualization**: Hands-on lab (for those at that stage).

**By the end of this module, you should be able to:**
- Apply effective `ggplot2` techniques, including appropriate geoms, scales, and coordinate systems.
- Analyze misleading visualizations and discuss better design choices.
- Use spatial data visualization techniques in the lab.

[Back to Menu](#2)

---

## Midsemester Check-In

You're halfway through the semester—great work so far! Your **midsemester check-in** is recommended by the end of this week. This is a chance to reflect on your progress and make any necessary adjustments.  

.pull-left[
### ✅ **What to Do:**  
- **Schedule Your Check-In**: Meet with me between **Weeks 5-8**.
- **Review Your Contract**: Check if you're meeting your commitments.  
- **Come prepared**: Think about what’s going well and what you might need help with.  
- **Request Contract Adjustments**: If needed, submit changes before the deadline.

[Back to Menu](#2)
]
.pull-right[
### 🔗 **Helpful Resources:**  
- **📑 Contract Repo**: [Access your contract](https://github.com/DataScience4Psych/2025_contracts/issues).  
- **📜 Contract Guidelines**: [Review the details](https://smasongarrison.github.io/syllabi/data-science.html#contract-details).  
- **📅 Check-In Schedule**: [Book your check-in](https://calendly.com/smasongarrison/checkin).  
```{r, echo=FALSE}
library(qrcode)
plot(qr_code('https://calendly.com/smasongarrison/checkin'))
```
]

[Back to Menu](#2)


---

## **Module 6: Confounding and Communication**

### **Today’s Plan:**
- **Understanding Confounding**: Case studies and discussion.
- **Simpson’s Paradox**: Exploring real-world implications.
- **Science Communication Strategies**: Effective storytelling with data.
- **Ugly Charts**: Hands-on lab (for those at that stage).

**By the end of this module, you should be able to:**
- Explain confounding and its impact on study design.
- Identify and analyze Simpson’s Paradox in real-world datasets.
- Apply principles of effective data storytelling.
- Critically evaluate misleading or poorly designed data visualizations.

[Back to Menu](#2)

---

## **Module 7: Web Scraping**

### **Today’s Plan:**
- **Introduction to Web Scraping**: Basics of extracting data from the web.
- **Techniques & Tools**: Using R for scraping, including `rvest` and `httr`.
- **Top 250 Movies on IMDB**: Exploring a real-world scraping example.
- **Hands-on Lab: Better Viz** (for those at that stage).

**By the end of this module, you should be able to:**
- Explain the fundamentals of web scraping and its applications.
- Use R packages like `rvest` and `httr` to scrape data.
- Apply CSS selectors using SelectorGadget.
- Work with real-world web scraping examples in R.

[Back to Menu](#2)
