---
title: "Visualizing categorical data<br> `r emo::ji('cat')`"
author: "S. Mason Garrison"
output:
  xaringan::moon_reader:
    css: "../slides.css"
    lib_dir: libs
    nature:
      ratio: "16:9"
      highlightLines: true
      highlightStyle: solarized-light
      countIncrementalSlides: false
      slideNumberFormat: ""
---

```{r child = "../setup.Rmd"}
```

```{r packages, echo=FALSE, message=FALSE, warning=FALSE}
library(tidyverse)
library(openintro)
loans_full_schema <- loans_full_schema %>%
  mutate(grade = factor(grade, ordered = TRUE))
```

class: middle

# Recap

---

## Variables

- **Numerical** variables can be classified as **continuous** or **discrete** based on whether or not the variable can take on an infinite number of values or only non-negative whole numbers, respectively.
- If the variable is **categorical**, we can determine if it is **ordinal** based on whether or not the levels have a natural ordering.

---

### Remember this Data?

```{r message=FALSE}
library(tidyverse)
starwars
```


---
### Perhaps now?
```{r}
glimpse(starwars)
```

---
#### Recode hair color

```{r }
starwars <- starwars %>%
  mutate(hair_color2 =
           fct_other(hair_color,
                     keep = c("black", "brown", "brown", "blond")
           )
  )
```
---


class: middle

# Bar plot

---

## Bar plot

```{r out.width="55%"}
ggplot(data = starwars, mapping = aes(x = gender)) +
  geom_bar()
```

---

## Segmented bar plot: counts

```{r out.width="55%"}

ggplot(data = starwars, mapping = aes(x = gender, 
                  fill = hair_color))+ #<<
  geom_bar()

```

---



## Segmented bar plots

```{r out.width="55%"}
ggplot(data = starwars, mapping = aes(x = gender, 
	   fill = hair_color2))+ #<<
  geom_bar()
```

---

## Segmented bar plots

```{r out.width="55%"}
ggplot(data = starwars, mapping = aes(x = gender, 
	   fill = hair_color2))+ #<<
  geom_bar()+ #<<
  coord_flip()
```

---

## Segmented bar plots: proportions

```{r out.width="45%"}
ggplot(data = starwars,
       mapping = aes(x = gender, fill = hair_color2)) +
  geom_bar(position = "fill") +
  coord_flip()
labs(y = "proportion")
```


---

.question[
    Which bar plot is a more useful representation for visualizing the relationship between gender and hair color?
  ]

.pull-left[
```{r echo=FALSE, out.width = "100%"}
ggplot(data = starwars, mapping = aes(x = gender, 
                  fill = hair_color2)) +
  geom_bar() +
  coord_flip()
```
]
.pull-right[
```{r echo=FALSE, out.width = "100%"}
ggplot(data = starwars, mapping = aes(x = gender, 
                  fill = hair_color2)) +
  geom_bar(position = "fill") +
  coord_flip() +
  labs(y = "proportion")
```
]

---

## Customizing bar plots

.pull-left[
```{r ref.label = "bar-custom", echo = FALSE, warning = FALSE}
```
]
.pull-right[.small[
```{r bar-custom, fig.show = "hide", warning = FALSE}
ggplot(starwars, aes(y = gender, #<<
                  fill = hair_color2)) +
  geom_bar(position = "fill") +
  labs( #<<
    x = "Proportion", #<<
    y = "Gender", #<<
    fill = "Hair Color", #<<
    title = "Hair Colors of Starwars", #<<
    subtitle = "by gender" #<<
  ) #<<
```
]



---

.your-turn[

    Time to actually play around with the Star Wars dataset!

      - Go to class git repo ([github.com/DataScience4Psych](https://github.com/DataScience4Psych)) and start `AE 03 - StarWars + Data visualization`.
    - Open the R Markdown document and complete the exercise (and if time allows, the stretch goal exercise).]

---


class: middle

# Relationships between numerical and categorical variables

---

## Already talked about...

- Coloring and faceting histograms and density plots
- Side-by-side box plots

---

## Violin plots

```{r warning = FALSE}
ggplot(loans, aes(x = homeownership, y = loan_amount)) +
  geom_violin()
```

---

## Ridge plots

```{r warning = FALSE}
library(ggridges)
ggplot(loans, aes(x = loan_amount, y = grade, fill = grade, color = grade)) + 
  geom_density_ridges(alpha = 0.5)
```

---


# Sources

- Mine Çetinkaya-Rundel's Data Science in a Box ([link](https://datasciencebox.org/))

---

class: middle

# Wrapping Up...
