---
title: "Title Line 1 <br> `r emo::ji('cat')`<br> Title Line 2"
author: "S. Mason Garrison"
output:
  xaringan::moon_reader:
    css: "../slides.css"
    lib_dir: libs
    self_contained: TRUE
    nature:
      ratio: "16:9"
      highlightLines: true
      highlightStyle: solarized-light
      countIncrementalSlides: false
#      slideNumberFormat: ""
---

```{r child = "../setup.Rmd"}
```


```{r packages, echo=FALSE, message=FALSE, warning=FALSE}
# Remember to compile
#xaringan::inf_mr(cast_from = "..")
#       slideNumberFormat: ""  
knitr::opts_chunk$set(knitr.duplicate.label = "allow")
library(tidyverse)
if (!require("emo")) devtools::install_github("hadley/emo")
library(emo)
```
---
class: middle

# Hello world!

---
# Welcome to [Course/Topic Name]

.br
`r emo::ji('cat')`

.large[
Welcome to our journey through [Course/Topic Name]. We'll explore together, learn, and grow.
]


---

# Course Overview Slide Template

## Course Overview

What will we cover today?

- **Introduction to [Topic]**
  - Key concepts and why it matters.
  
- **Deep Dive: [Specific Topic]**
  - Exploring in detail.
  
- **Practical Application**
  - How can you apply this knowledge?

- **Q&A**
  - Your questions answered.

---

Interactive Question Slide Template

## Let's Discuss

- What is your understanding of [Topic/Question]?

--

- How do you see [Topic/Question] applying to your work/life?

.large[
Feel free to share your thoughts and questions.
]


# Topic Title Here

.small[
.instructions[
Go to [your_instruction_link](https://yourlinkhere.com/), and perform the action (unless you already have done so). 
]
]

Tips for __________:<sup>✦</sup>

.small[
- Tip 1 with your content.
- Tip 2 with your content.
- Tip 3 with your content.
- More tips as needed.
]

--
.small[ 
<sup>✦</sup> Source: [Your Source Name](http://yoursourceurl.com) by Author/Creator Name
]

---

class: middle

# Wrapping Up...


