---
title: "Code and Data Slides Template"
author: "S. Mason Garrison"
output:
  xaringan::moon_reader:
    css: "../slides.css"
    lib_dir: libs
    nature:
      ratio: "16:9"
      highlightLines: true
      highlightStyle: solarized-light
      countIncrementalSlides: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(knitr)
library(palmerpenguins)
```

# Simple Code Chunk

Here's a simple R code chunk:

```{r simple-code}
x <- 3 * 4
y <- 10 - 2
z <- x + y
print(z)
```

---

# Code with Output

This code chunk will show both the code and its output:

```{r code-with-output}
# Create a vector
numbers <- c(1, 2, 3, 4, 5)

# Calculate mean and standard deviation
mean_value <- mean(numbers)
sd_value <- sd(numbers)

# Print results
cat("Mean:", mean_value, "\n")
cat("Standard Deviation:", sd_value)
```

---

# Code with Highlighted Lines

In this example, we'll highlight important lines:

```{r highlight-lines, highlight.output=c(1, 3)}
# Load the tidyverse package
library(tidyverse)

# Create a simple dataframe
df <- tibble(
  x = 1:5,
  y = c(2, 4, 6, 8, 10)
)

# Plot the data
ggplot(df, aes(x, y)) +
  geom_point() +
  geom_line()
```

---

# Data Visualization Code

Here's an example of creating a data visualization with code:

```{r data-viz, fig.height=4}
ggplot(penguins, aes(x = flipper_length_mm, y = body_mass_g, color = species)) +
  geom_point(alpha = 0.7) +
  labs(title = "Penguin Size",
       x = "Flipper Length (mm)",
       y = "Body Mass (g)",
       color = "Species") +
  theme_minimal()
```

---

# Data Manipulation

This slide demonstrates data manipulation with dplyr:

```{r data-manip}
penguins %>%
  group_by(species, island) %>%
  summarize(
    mean_body_mass = mean(body_mass_g, na.rm = TRUE),
    mean_flipper_length = mean(flipper_length_mm, na.rm = TRUE)
  ) %>%
  arrange(desc(mean_body_mass))
```

---

# Interactive Code Output

This slide shows how to create an interactive table:

```{r interactive-table}
library(DT)

penguins %>%
  select(species, island, body_mass_g, flipper_length_mm) %>%
  datatable()
```

---

# Code and Plot Side by Side

.pull-left[
```{r side-by-side-code, fig.show = "hide"}
ggplot(mtcars, aes(x = wt, y = mpg)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  labs(
    title = "Car Weight vs. Miles Per Gallon",
    x = "Weight (1000 lbs)",
    y = "Miles Per Gallon"
  )
```
]
.pull-right[
```{r ref.label="side-by-side-code", echo = FALSE, out.width = "100%"}
```
]
```

---

# Importing Data

Here's how you might import and display a dataset:

```{r import-data}
# Read CSV file (replace with your actual file path)
df <- read_csv("path/to/your/data.csv")

# Display the first few rows
head(df)
```

---

# Statistical Analysis

This slide demonstrates a simple statistical test:

```{r stat-analysis}
# Perform a t-test
t_test_result <- t.test(flipper_length_mm ~ species, data = penguins)

# Print the results
print(t_test_result)
```

---

## Residuals Plot

.pull-left[
```{r residuals-plot, fig.show="hide"}
ht_wt_fit <- linear_reg() %>%
  set_engine("lm") %>%
  fit(Height_in ~ Width_in, data = pp)

ht_wt_fit_aug <- augment(ht_wt_fit$fit)

ggplot(ht_wt_fit_aug, mapping = aes(x = .fitted, y = .resid)) +
  geom_point(alpha = 0.5) +
  geom_hline(yintercept = 0, color = "#bf2986ff", lty = "dashed") +
  labs(x = "Predicted Height", y = "Residuals")
```
]
.pull-right[
```{r ref.label="residuals-plot", echo=FALSE, out.width="100%"}
```
]


---


## Model Output

```{r model-output}
linear_reg() %>%
  set_engine("lm") %>%
  fit(Height_in ~ Width_in, data = pp) %>%
  tidy()
```

---


## Data Exploration

.pull-left[
```{r data-explore, fig.show="hide"}
ggplot(pp, aes(x = Width_in)) +
  geom_histogram(binwidth = 5) +
  labs(x = "Width (inches)", y = NULL)
```
]
.pull-right[
```{r ref.label="data-explore", echo=FALSE, out.width="100%"}
```
]
---




## Cross Validation Metrics

```{r cv-metrics}
collect_metrics(office_fit_rs)
```


---

## Bootstrap Distribution

```{r boot-dist, fig.height=4}
ggplot(boot_df, aes(x = stat)) +
  geom_histogram(binwidth = 100) +
  labs(title = "Bootstrap distribution of means")
```
