---
title: "Conclusion Slides Template"
author: "S. Mason Garrison"
output:
  xaringan::moon_reader:
    css: "../slides.css"
    lib_dir: libs
    nature:
      ratio: "16:9"
      highlightLines: true
      highlightStyle: solarized-light
      countIncrementalSlides: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(tidyverse)
library(emo)
```

class: middle

# Summary

Key points from today's presentation:

1. First main point
2. Second main point
3. Third main point
4. Fourth main point

---

# What We've Learned

.pull-left[
## Concepts
- Concept A
- Concept B
- Concept C
]

.pull-right[
## Skills
- Skill X
- Skill Y
- Skill Z
]

---

# Conclusion

.large[
In conclusion, we've explored [main topic]. Remember:

- Key takeaway 1
- Key takeaway 2
- Key takeaway 3

These insights will help you [achieve some goal or understand some concept better].
]

---

# Next Steps

Moving forward:

1. Apply what you've learned to [specific task or project]
2. Explore further resources on [related topic]
3. Practice [particular skill] regularly
4. Prepare for our next session on [upcoming topic]

---

class: middle

# Wrapping Up... 

.hand[
Thank you for your attention and participation!
]

---

# Q&A

.large[
Any final questions?
]

---

# Contact Information

- Email: _garrissm@wfu.edu_
- Website: [smasongarrison.com](https://www.smasongarrison.com)
- Twitter: [@smasongarrison](https://twitter.com/smasongarrison)
- GitHub: [github.com/smasongarrison](https://github.com/smasongarrison)
- YouTube: [youtube.com/smasongarrison](https://youtube.com/smasongarrison)

```{r qrcode, echo=FALSE, fig.align = "center", out.width = "30%", caption="Scan the QR code to visit the Data Science for Psych website"}
library(qrcode)
code <- qr_code("https://DataScience4Psych.github.io/DataScience4Psych/")
plot(code)
```

---

class: center, middle

# Thank You!

